Implementation of poloidal density variation in collision operator

### 2019-03-05 doc/manual/version3/overview.tex
Working on manual
Substantial revision to manual

The kinetic equations solved in \sfincs~have four independent variables: poloidal angle $\theta$, toroidal angle $\zeta$,
normalized speed $x = v / v_{thermal}$, and pitch angle $\xi = v_{||}/v$.  The third velocity coordinate (gyro-angle) does not appear
since gyro-averaged equations are solved.  The flux surface label (radius) coordinate is only a parameter, rather than a full independent variable,
since a radially local approximation is made.

@@

@@

The output from a single \sfincs~computation is saved in \HDF~format in the file {\ttfamily sfincsOutput.h5}.
To browse this file you can enter {\ttfamily h5dump sfincsOutput.h5|less} from the command line.
Every array saved in this file is annotated with strings that describe the array dimensions,
for example \Ntheta$\times$ \Nzeta. One of the array dimensions may be {\ttfamily iteration},
which can either indicate the iteration of the Newton solver for a nonlinear calculation,
or which right-hand side vector was used when computing a transport matrix.
Many of the variables in the output file are also annotated with
text that describes their meaning and normalization.

>>>>

The kinetic equations solved in \sfincs~have four independent variables: poloidal angle $\theta$, toroidal angle $\zeta$,
normalized speed $x = v / v_{thermal}$, and pitch angle $\xi = v_{||}/v$.  The third velocity coordinate (gyro-angle) does not appear
since gyro-averaged equations are solved.  The flux surface label (radius) coordinate is only a parameter, rather than a full independent variable,
since a radially local approximation is made. 
The solution is the first-order distribution function $f_{s1} \left(\theta, \zeta, x, \xi\right)$ (where the full distribution is $f_s = f_{s0} + f_{s1}$) on the flux surface for all species $s$, and if quasi-neutrality is solved also the electrostatic potential variation on the flux surface $\Phi_1 \left(\theta, \zeta\right) = \Phi-\left<\Phi\right>$.

@@

\item 
The radial classical particle and heat fluxes are calculated.

@@

The output from a single \sfincs~computation is saved in \HDF~format in the file %{\ttfamily sfincsOutput.h5}.
specified by \parlink{outputFileName}.
To browse this file you can enter {\ttfamily h5dump outputFileName|less} %{\ttfamily h5dump sfincsOutput.h5|less} 
from the command line.
Every array saved in this file is annotated with strings that describe the array dimensions,
for example \Ntheta$\times$ \Nzeta. One of the array dimensions may be {\ttfamily iteration},
which can either indicate the iteration of the Newton solver for a nonlinear calculation,
or which right-hand side vector was used when computing a transport matrix.
Many of the variables in the output file are also annotated with
text that describes their meaning and normalization. 
Some of the most frequently used output quantities are detailed in chapter~\ref{ch:output}.

### 2019-03-05 doc/manual/version3/runs.tex
Substantial revision to manual

You can choose any reference parameters you like, not just the values
suggested here. However, if you use a {\ttfamily vmec} or {\ttfamily .bc} magnetic equilibrium
by choosing \parlink{geometryScheme} = 5, 11, or 12, then you MUST use $\bar{B}$ = 1 Tesla and $\bar{R}$ = 1 meter.
The code ``knows'' about the reference values only through
the 3 combinations \parlink{Delta}, \parlink{alpha}, and \parlink{nu\_n}
in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.

@@

0: Specify the flux surface using \parlink{psiHat\_wish}.\\

1: Specify the flux surface using \parlink{psiN\_wish}.\\

2: Specify the flux surface using \parlink{rHat\_wish}.\\

3: Specify the flux surface using \parlink{rN\_wish}.\\

@@

%\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
\section{Nonlinear calculations, includePhi1 = {\ttfamily .true.}}
\label{sec:qn}

One choice you should consider in setting up a computation is whether or not
to include variation on the flux surface of the electrostatic potential, $\Phi_1(\theta,\zeta)$.
Such variation does occur to some degree in a real plasma, but it is 
neglected in analytical theory and in many codes such as \dkes.  It can be proved
that including $\Phi_1$ has no effect on the particle or heat fluxes, parallel flows, or bootstrap current
when $E_r=0$, but generally there can be some difference when $E_r \ne 0$.
(There is a subtlety in showing that the heat flux is the same with and without $\Phi_1$,
discussed in the notes 20150325-01 in \path{sfincs/doc/}.)

In \sfincs, you can choose whether or not to
include $\Phi_1$ using the parameter {\ttfamily \hyperlink{includePhi1}{includePhi1}} in the 
{\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}}
namelist. 
Furthermore, the parameters \parlink{includePhi1inKineticEquation}, \parlink{includePhi1InCollisionOperator} and \parlink{quasineutralityOption} can be used to control how $\Phi_1$ is included in the system of equations. 
If and only if $\Phi_1$ is included, a quasineutrality equation is solved
at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
(quasineutrality), the system matrix is slightly larger when \parlink{includePhi1} is \true.
Specifically, the number of rows and columns are each increased by \Ntheta$\times$\Nzeta$+1$.  This increase is miniscule compared
to the number of rows and columns associated with the kinetic equation, which depends not only on real space
but also on velocity space and species.  
%Thus, there is very little extra computational cost associated
%with \parlink{includePhi1}. 
Thus, there is typically not a substantial increase in memory requirements associated with \parlink{includePhi1} (however, setting the preconditioner option \parlink{reusePreconditioner} = \false~can lead to a significant increase in memory requirements). 
The time requirements will normally increase when \parlink{includePhi1} is \true~because the system of equations is then nonlinear. 
The nonlinear equations are solved with a Newton method using the Scalable Nonlinear Equations Solvers (SNES) in \PETSc. SNES internally employs KSP for the solution of its linear systems. 
Note that for a nonlinear calculation the \sfincs~output file \parlink{outputFileName} will contain the results of each nonlinear iteration in the relevant output parameters. The final result of the calculation is stored in the last position of the arrays. 
The resolution parameters required for numerical convergence in nonlinear calculations are often similar to the parameters in the corresponding linear calculations (based on experience). 
A sanity check to control that the nonlinear calculation is not incorrect (however it is not a guarantee for numerical convergence), is to check that the total flux-surface averaged particle flux for the lowest order distribution function $\displaystyle \left\langle \int d^3 v \, f_{s0} \left(\bm{v}_{\mathrm{d} s} + \bm{v}_{E}\right) \cdot \nabla \psi \right\rangle$ of each plasma species $s$ comes out as 0. This can e.g. be done by comparing {\ttfamily particleFlux\_vm0\_psiHat} to {\ttfamily particleFlux\_vE0\_psiHat} in the \sfincs~output file \parlink{outputFileName} and see that they are equal but with opposite sign.

@@

You can choose to either include or not include the poloidal and toroidal magnetic drifts.
These drifts are turned off by default.  To turn them on, all you need to do is
set \parlink{magneticDriftScheme} = 1 in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
(Setting \parlink{magneticDriftScheme} = 2 uses a slightly different parallel magnetic drift, which gives
indistinguishable results to setting 1 for all cases examined so far, and which is in fact exactly identical to setting 1
in the limit of vanishing plasma beta.)
If the poloidal/toroidal magnetic drifts are turned on, you must use VMEC
geometry (\parlink{geometryScheme} = 5), since the magnetic drifts
depend on various derivatives of the components of the magnetic field which
are not available in the simplified geometry models.

@@

If the electrostatic potential is not constant on flux surfaces and poloidal/toroidal magnetic drifts
are included, certain terms exist in the kinetic equation which have not yet been implemented in the code.
Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme} $> 0$ 
and  \parlink{includePhi1} = \true.

@@

By setting \parlink{RHSMode} = 3, \sfincs~can be run in a mode
where it solves the same kinetic equation (prior to discretization) as 
\dkes~and other monoenergetic codes.
When \parlink{RHSMode} = 3, the values of \parlink{Zs}, \parlink{THats}, \parlink{nHats},
\parlink{mHats}, \parlink{nu\_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
Instead, the collisionality is set by \parlink{nuPrime}, and the radial electric field is set
by \parlink{EStar}.  The first of these quantities is the dimensionless collisionality
\begin{equation}
\mbox{\parlink{nuPrime}} = \frac{(G+\iota I) \nu}{v B_0}
\end{equation}
where $G$ and $I$ are defined in (\ref{eq:covariant}), $\iota=1/q$ is the rotational transform,
$v$ is the speed at which the monoenergetic calculation is being performed, and $B_0$ is the (0,0) Fourier harmonic of $B$
with respect to the Boozer poloidal and toroidal angles. The collision
frequency $\nu$ is here the value of $\nu_\mathrm{ii}$ one would have if
$v$ were the thermal speed. That is, in SI units,
\begin{equation}
  \nu=\frac{4}{3\sqrt{\pi}}\frac{n Z^4e^4\ln \Lambda}{4\pi\epsilon_0^2m^2v^3}.
\end{equation}
%
The normalized radial electric field is
\begin{equation}
\mbox{\parlink{EStar}} = \frac{cG}{\iota v B_0} \frac{d\Phi}{d\psi}
\end{equation}
(Gaussian units).
When {\ttfamily RHSMode} == 1, {\ttfamily nuPrime} and {\ttfamily EStar} are ignored.
\todo{Should be change the behavior of RHSMode=2 so it uses nuPrime and EStar instead of nu\_n?}

>>>>

You can choose any reference parameters you like, not just the values
suggested here. However, if you use a {\ttfamily vmec} or {\ttfamily .bc} magnetic equilibrium
by choosing \parlink{geometryScheme} = 5, 11, or 12, then you MUST use $\bar{B}$ = 1 Tesla and $\bar{R}$ = 1 meter.
The code ``knows'' about the reference values only through
the 3 combinations \parlink{Delta}, \parlink{alpha}, and \underscoreparlink{nu\_n}{nu_n}
in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.

@@

0: Specify the flux surface using \underscoreparlink{psiHat\_wish}{psiHat_wish}.\\

1: Specify the flux surface using \underscoreparlink{psiN\_wish}{psiN_wish}.\\

2: Specify the flux surface using \underscoreparlink{rHat\_wish}{rHat_wish}.\\

3: Specify the flux surface using \underscoreparlink{rN\_wish}{rN_wish}.\\

@@

%\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
\section{Calculations with $\Phi_1(\theta,\zeta)$, includePhi1 = {\ttfamily .true.}}
\label{sec:qn}

One choice you should consider in setting up a computation is whether or not
to include variation on the flux surface of the electrostatic potential, $\Phi_1(\theta,\zeta)$.
Such variation does occur to some degree in a real plasma, but it has traditionally been  
neglected in analytical theory and in many codes such as \dkes~(however, nowadays analytical modelling exists, see e.g. \cite{Buller2018} and references therein).  It can be proved
that including $\Phi_1$ has no effect on the particle or heat fluxes, parallel flows, or bootstrap current
when $E_r=0$, but generally there can be some difference when $E_r \ne 0$.
(There is a subtlety in showing that the heat flux is the same with and without $\Phi_1$,
discussed in the notes 20150325-01 in \path{sfincs/doc/}.)

In \sfincs, you can choose whether or not to
include $\Phi_1$ using the parameter \parlink{includePhi1} in the 
{\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}}
namelist. 
Furthermore, the parameters \parlink{includePhi1InKineticEquation}, \parlink{includePhi1InCollisionOperator} and \parlink{quasineutralityOption} can be used to control how $\Phi_1$ is included in the system of equations. 
\sfincs~also allows to read $\Phi_1$ as an input by using \parlink{readExternalPhi1}. This is useful for looking at the effect on the transport by a $\Phi_1$ generated by some mechanism not treated by \sfincs, or for using $\Phi_1$ calculated by another code. Moreover, with $\Phi_1$ as an input the runs are typically cheaper than if \sfincs~calculates $\Phi_1$ itself, so if the potential variation is not expected to change in a parameter scan this setting could be useful to reuse the same $\Phi_1$ and decrease the computational costs. 
If and only if $\Phi_1$ is included as an unknown (i.e. not if $\Phi_1$ is an input with \parlink{readExternalPhi1}), a quasineutrality equation is solved
at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
(quasineutrality), the system matrix is slightly larger when \parlink{includePhi1} is \true.
Specifically, the number of rows and columns are each increased by \Ntheta$\times$\Nzeta$+1$.  This increase is miniscule compared
to the number of rows and columns associated with the kinetic equation, which depends not only on real space
but also on velocity space and species.  
%Thus, there is very little extra computational cost associated
%with \parlink{includePhi1}. 
Thus, there is typically not a substantial increase in memory requirements associated with \parlink{includePhi1} (however, setting the preconditioner option \parlink{reusePreconditioner} = \false~can lead to a significant increase in memory requirements). 
The time requirements will normally increase when \parlink{includePhi1} is \true~because the system of equations is then nonlinear (except if \parlink{readExternalPhi1} is \true). 
The nonlinear equations are solved with a Newton method using the Scalable Nonlinear Equations Solvers (SNES) in \PETSc. SNES internally employs KSP for the solution of its linear systems. 
Note that for a nonlinear calculation the \sfincs~output file \parlink{outputFileName} will contain the results of each nonlinear iteration in the relevant output parameters. The final result of the calculation is stored in the last position of the arrays. 
The resolution parameters required for numerical convergence in nonlinear calculations are often similar to the parameters in the corresponding linear calculations (based on experience). 
A sanity check to control that the nonlinear calculation is not incorrect (however it is not a guarantee for numerical convergence), is to check that the total flux-surface averaged particle flux for the lowest order distribution function $\displaystyle \left\langle \int d^3 v \, f_{s0} \, \bm{v}_{\mathrm{d} s} \cdot \nabla \psi \right\rangle = \left\langle \int d^3 v \, f_{s0} \left(\bm{v}_{\mathrm{m} s} + \bm{v}_{E}\right) \cdot \nabla \psi \right\rangle$ of each plasma species $s$ comes out as 0. This can e.g. be done by comparing \underscoreparlink{particleFlux\_vm0\_psiHat}{particleFlux_vm0_psiHat} to \underscoreparlink{particleFlux\_vE0\_psiHat}{particleFlux_vE0_psiHat} in the \sfincs~output file \parlink{outputFileName} and see that they are equal but with opposite sign.

@@

You can choose to either include or not include the poloidal and toroidal magnetic drifts.
These drifts are turned off by default.  To turn them on, all you need to do is
set \parlink{magneticDriftScheme}~$>$~0 (with the default being \parlink{magneticDriftScheme}~=~1) in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist. 
(Setting \parlink{magneticDriftScheme}~=~2 uses a slightly different parallel magnetic drift, which gives
indistinguishable results to setting 1 for all cases examined so far, and which is in fact exactly identical to setting 1
in the limit of vanishing plasma beta.)
If the poloidal/toroidal magnetic drifts are turned on, you must use VMEC
geometry (\parlink{geometryScheme}~=~5), since the magnetic drifts
depend on various derivatives of the components of the magnetic field which
are not available in the simplified geometry models. 
(In fact, to include poloidal/toroidal magnetic drifts in a radially local code like \sfincs~is somewhat ambiguous since the approximation made when neglecting the radial derivatives of the unknowns is different in different coordinate system.)

@@

If the electrostatic potential is not constant on flux surfaces and poloidal/toroidal magnetic drifts
are included, certain cross-terms exist in the kinetic equation which have not yet been implemented in \sfincs.
Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme}~$>$~0 
and  \parlink{includePhi1} = \true, but it is allowed for testing purposes.

@@

By setting \parlink{RHSMode} = 3, \sfincs~can be run in a mode
where it solves the same kinetic equation (prior to discretization) as 
\dkes~and other monoenergetic codes.
When \parlink{RHSMode} = 3, the values of \parlink{Zs}, \parlink{THats}, \parlink{nHats},
\parlink{mHats}, \underscoreparlink{nu\_n}{nu_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
Instead, the collisionality is set by \parlink{nuPrime}, and the radial electric field is set
by \parlink{EStar}.  The first of these quantities is the dimensionless collisionality
\begin{equation}
\mbox{\parlink{nuPrime}} = \frac{(G+\iota I) \nu}{v B_0}
\end{equation}
where $G$ and $I$ are defined in (\ref{eq:covariant}), $\iota=1/q$ is the rotational transform,
$v$ is the speed at which the monoenergetic calculation is being performed, and $B_0$ is the (0,0) Fourier harmonic of $B$
with respect to the Boozer poloidal and toroidal angles. The collision
frequency $\nu$ is here the value of $\nu_\mathrm{ii}$ one would have if
$v$ were the thermal speed. That is, in SI units,
\begin{equation}
  \nu=\frac{4}{3\sqrt{\pi}}\frac{n Z^4e^4\ln \Lambda}{4\pi\epsilon_0^2m^2v^3}.
\end{equation}
%
The normalized radial electric field is
\begin{equation}
\mbox{\parlink{EStar}} = \frac{cG}{\iota v B_0} \frac{d\Phi}{d\psi}
\end{equation}
(Gaussian units).
When {\ttfamily RHSMode} == 1, {\ttfamily nuPrime} and {\ttfamily EStar} are ignored.
\todo{Should be change the behavior of RHSMode=2 so it uses nuPrime and EStar instead of nu\_n?}
